Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-15 main.Rmd
added some missing parts

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM. ^[Ruth, what about the stuff you did on the cluster with the GPUs? This isn't correct, is it?].

@@

Hyperparameter scans were performed in ^[Ruth, can you outline the packages and methods in scikit-learn/tensorflow that you used for this] Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). All modle training was performed as described in section Model Training.

@@

Fine tuning was performed manually using Tensorflow/scikit learn ^[Ruth, can you add the software and versions your scripts used here] Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).

@@

To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019]. We extracted the convolutional layer `convd_1` and computed the gradient of the model output with respect to the output of `convd_1` The feature map was weighted by every channel and mean result taken. All activation maps were summed, normalised and smoothed using Fourier analysis ^[Ruth, there need to be references, mentions of software and versions used, it isn't easy to see what you actually did from this description]. The code we used for computing and visualising these heatmaps can be found at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .

@@

We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding. To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 120 * 50` models in total. ^[Ruth, can you confirm that this interpretation is correct]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam) ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction, the best performing values can be see in Table \@ref(tab:possiblecombination). From this we had twelve candidate models to develop further.

@@

We then manually ran and checked the accuracy and loss of the twelve models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. Parameters varied in this phase are listed in \@ref(tab:tunedhyperparam). ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam). The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal). We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.

@@

Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[Ruth, have we any idea what this fraction might be] of the possible hyperparameter and parameter space was explored we still compared 6000 models. At a run time of around ^[Ruth, approx run time per model] So even for relatively small data sets, a large amount of specialised GPU compute power is required.

@@

```{r, fixedhyperparam}
apa_table(
  readr::read_tsv("tables/fixed_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
)
```

```{r, tunedhyperparam}
apa_table(
  readr::read_tsv("tables/tuned_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
)
```

>>>>

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by hyperparamter scans. All models were implemented in Python 3.6.9 using the deep learning API Keras 2.2.4 with Tensorflow 1.12.0 backend, using GPU NVIDIA GK110GL Tesla K20c and CPU AMD Opteron(TM) Processor 6272 with memory 128 GB.

@@

Hyperparameter scans were performed using random search, a hyperparameter optimization method which 
implements a randomized search over hyperparameters, where each hyperparameter setting is randomly sampled from a distribution over possible hyperparameter values [@Bergstra2012]. In particularly, we used RandomSearchCV(), the implementation of random search method in `scikit-learn 0.19.2` [@scikit-learn] together with KerasClassifier() which is implementation of the scikit-learn classifier API for keras. Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). All model training was performed as described in section Model Training.

@@

Fine tuning was performed manually using `keras 2.2.4` together with metrics module and KFold cross validator from `scikit-learn 0.19.2`. Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).

@@

To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we adapted a Grad-CAM (Gradient-weighted Class Activation Mapping) approach [@Selvaraju2019]. We extracted the convolutional layer `convd_1` and computed the gradient of the model output with respect to the output of `convd_1`. The feature map was weighted by every channel and mean result taken. All activation maps were summed, normalised and smoothed using Fourier analysis. We used `tensorflow 1.12.0` to compute the activation maps and discrete fourier transform (numpy.fft) from `numpy 1.17.3` to smooth the result. The code we used for computing and visualising these heatmaps can be found at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/heatmaps) .

@@

We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding. To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 3 * 4 * 50 * 5` models in total. ^[For each architecture, 50 models are selected randomly. Therefore there are 3 taxon * 4 architecture * 50 models * 5 cross validation]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam). For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction. ^[\@ref(tab:possiblecombination) is table of total number of all hyperparameter settings in the whole hyperparameter space for each architecture] From this we had twelve candidate models to develop further.

@@

We then manually ran and checked the accuracy and loss of the twelve models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. Parameters varied in this phase are listed in \@ref(tab:tunedhyperparam). Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam). The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal). We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.

@@

Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[This fraction depends for every model, the parameter space is shown in table \@ref(tab:possiblecombination)] of the possible hyperparameter and parameter space was explored we still compared `r 3 * 4 * 50 * 5` models. At a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding. So even for relatively small data sets, a large amount of specialised GPU compute power is required.

@@

```{r, fixedhyperparam}
# apa_table(
#   readr::read_tsv("tables/fixed_hyperparam.tab"),
#   digits = 0,
#   align = 'c',
#   caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
# )

readr::read_tsv("tables/fixed_hyperparam.tab") %>%
  knitr::kable(
    "latex", 
    longtable = T, 
    booktabs = T,
    digits = 0,
    align = 'c',
    caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
  ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
```

```{r, tunedhyperparam}
# apa_table(
#   readr::read_tsv("tables/tuned_hyperparam.tab"),
#   digits = 0,
#   align = 'c',
#   caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
# )

readr::read_tsv("tables/tuned_hyperparam.tab") %>%
  knitr::kable(
    "latex", 
    longtable = T, 
    booktabs = T,
    digits = 0,
    align = 'c',
    caption = 'Hyperparameters tuned in manual scanning of hyperparameter space for each model'
  ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
```
